---
title: "Figure Fig2H — Human EGL surface area (Term vs Preterm), n = 4 per group"
subtitle: "Mann–Whitney U (two-tailed); style matches Fig 2 B–F"
author: "Kratimenos Lab"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 2.6, fig.height = 4.4, dpi = 300)
library(readr); library(dplyr); library(ggplot2); library(knitr); library(kableExtra)
```

# Overview

Panel **Fig2H** (Human EGL surface area (Term vs Preterm)) — n = 4 per group, matched on GA from the larger
Prism cohort. Values were recovered from the original published Prism panel
(per-subject points) so this Rmd exactly reproduces the published bars, error
bars and significance call without going through Prism.

Statistical test: **two-tailed Mann–Whitney U** (as in the source n = 4
analysis). Visual style aligned with Fig 2 B–F.

# Data

```{r load}
group_levels <- c("Term", "Preterm")
dat <- read_csv("data/Fig2H_n4_data.csv", show_col_types = FALSE) |>
  mutate(Group = factor(Group, levels = group_levels))

dat |> group_by(Group) |>
  summarise(n = n(), mean = mean(Value), sd = sd(Value),
            median = median(Value), min = min(Value), max = max(Value),
            .groups = "drop") |>
  kable(digits = 3, caption = "Per-group summary") |>
  kable_styling(full_width = FALSE)
```

# Statistical test

```{r test}
mw <- wilcox.test(Value ~ Group, data = dat, exact = FALSE, correct = FALSE)
p_value <- mw$p.value
broom::tidy(mw) |>
  kable(digits = 4, caption = "Mann–Whitney U (two-tailed)") |>
  kable_styling(full_width = FALSE)
```

# Figure

```{r plot, fig.width=2.6, fig.height=4.4}
fill_cols <- c("Term" = "#F1E3CA", "Preterm" = "#8FA3BC")

summ <- dat |> group_by(Group) |>
  summarise(mean = mean(Value), sd = sd(Value),
            top  = max(Value) + sd(Value), .groups = "drop")

sig_lab <- dplyr::case_when(
  p_value < 0.0001 ~ "****",
  p_value < 0.001  ~ "***",
  p_value < 0.01   ~ "**",
  p_value < 0.05   ~ "*",
  TRUE             ~ "ns")

span <- 45; step <- span * 0.065
base <- max(summ$top) + step * 0.7
y_top <- 45
if (sig_lab != "ns") y_top <- max(y_top, base + step)

set.seed(7)
p <- ggplot(summ, aes(Group, mean)) +
  geom_col(aes(fill = Group), width = 0.62, colour = "black", linewidth = 0.35) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.22, linewidth = 0.35) +
  geom_point(data = dat, aes(Group, Value, fill = Group),
             position = position_jitter(width = 0.12, seed = 7),
             size = 1.6, shape = 21, colour = "black", stroke = 0.3,
             show.legend = FALSE) +
  scale_fill_manual(values = fill_cols) +
  scale_y_continuous(limits = c(0, y_top),
                     breaks = seq(0, 45, 10), expand = c(0, 0)) +
  labs(title = NULL, x = NULL, y = "EGL surface area (µm²)") +
  theme_classic(base_size = 12, base_family = "Arial") +
  theme(legend.position  = "none",
        plot.title       = element_text(hjust = 0.5),
        axis.text.x      = element_text(angle = 90, hjust = 1, vjust = 0.5),
        plot.background  = element_rect(fill = "white", color = NA),
        panel.background = element_rect(fill = "white", color = NA))

if (sig_lab != "ns") {
  yy <- base
  p <- p +
    annotate("segment", x = 1, xend = 2, y = yy, yend = yy, linewidth = 0.3) +
    annotate("segment", x = 1, xend = 1, y = yy, yend = yy - step * 0.18, linewidth = 0.3) +
    annotate("segment", x = 2, xend = 2, y = yy, yend = yy - step * 0.18, linewidth = 0.3) +
    annotate("text", x = 1.5, y = yy + step * 0.05, label = sig_lab,
             vjust = 0, fontface = "bold", size = 4.4)
}
p

ggsave("Fig2H.pdf", p, width = 2.6, height = 4.4, device = cairo_pdf)
ggsave("Fig2H.png", p, width = 2.6, height = 4.4, dpi = 300, bg = "white")
```

# Verification vs source (n = 4 cohort)

| Statistic | Source | This analysis |
|---|---|---|
| Term n / mean / SD | 4 / 18.375 / 4.785 | `r dat |> filter(Group=="Term") |> nrow()` / `r round(mean(dat$Value[dat$Group=="Term"]), 3)` / `r round(sd(dat$Value[dat$Group=="Term"]), 3)` |
| Preterm n / mean / SD | 4 / 31.378 / 5.214 | `r dat |> filter(Group=="Preterm") |> nrow()` / `r round(mean(dat$Value[dat$Group=="Preterm"]), 3)` / `r round(sd(dat$Value[dat$Group=="Preterm"]), 3)` |
| Test | Mann–Whitney U (two-tailed) | p = `r format(p_value, digits = 3)` (`r sig_lab`) |

# Session info

```{r session, echo=FALSE}
sessionInfo()
```

