---
title: "Figure 4N — Mitochondrial Health Pie Charts (Granule Cells)"
author: "Kratimenos Lab"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: hide
    df_print: paged
params:
  plot_order: !r c(3, 4, 5, 1, 2)
  start_deg: 0
  direction: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
suppressPackageStartupMessages({
  library(readr); library(dplyr); library(tidyr); library(ggplot2)
  library(patchwork); library(cowplot); library(grid); library(knitr)
})

CELL <- "GCs"   # "GCs" or "PCs"
DATA_FILE <- "data/Fig4NO_counts.csv"
```

# Overview

**Figure 4N** — mitochondrial-health distribution in granule cells
across 4 experimental groups (Sal-N, Sal-Hx, LPS-N, LPS-Hx) x 2 ages
(P11, P45), scored into 5 EM-morphology categories.

| Code | Category                                            | Colour      |
|------|-----------------------------------------------------|-------------|
| 1    | Healthy matrix, aligned cristae                     | medium blue |
| 2    | Healthy matrix, disorganized cristae                | light blue  |
| 3    | Poor matrix, disorganized cristae                   | pale yellow |
| 4    | Small vacuoles                                      | orange      |
| 5    | Mitophagy, minimal cristae, large vacuoles          | red         |

# Load data

```{r load}
counts_wide <- read_csv(DATA_FILE, show_col_types = FALSE)
dat <- counts_wide |>
  filter(cell_type == CELL) |>
  pivot_longer(cols = starts_with("cat"),
               names_to = "category", values_to = "count") |>
  mutate(category = sub("^cat", "", category)) |>
  mutate(
    condition = factor(condition, levels = c("Sal-N","Sal-Hx","LPS-N","LPS-Hx")),
    age       = factor(age,       levels = c("P11", "P45")),
    # factor-level order = plot order around the pie (CW from start)
    category  = factor(category,  levels = as.character(params$plot_order))
  ) |>
  arrange(condition, age, category)

dat |>
  mutate(category = paste0("cat", as.character(category))) |>
  pivot_wider(names_from = category, values_from = count) |>
  kable(caption = sprintf("Per-condition mitochondrial-health counts (%s)", CELL))
```

# Plot

```{r palette}
# Office / PowerPoint default chart palette — matches the published figure.
cat_colors <- c(
  "1" = "#4472C4",   # medium blue
  "2" = "#B4C7E7",   # light blue
  "3" = "#FFF2CC",   # pale yellow
  "4" = "#ED7D31",   # orange
  "5" = "#C00000"    # brick red
)
start_rad <- params$start_deg * pi / 180
pie_dir   <- params$direction
```

```{r plot-helpers}
pie_one <- function(df_row) {
  ggplot(df_row, aes(x = "", y = count, fill = category)) +
    geom_col(width = 1, colour = "white", linewidth = 0.5,
             position = position_stack(reverse = TRUE)) +
    coord_polar(theta = "y", start = start_rad, direction = pie_dir) +
    scale_fill_manual(values = cat_colors, drop = FALSE, guide = "none") +
    theme_void(base_family = "Arial") +
    theme(plot.margin = margin(1, 1, 1, 1))
}

condition_header <- function(label) {
  ggplot() +
    annotate("segment", x = 0.05, xend = 0.10, y = 0.10, yend = 0.55, linewidth = 0.4) +
    annotate("segment", x = 0.10, xend = 0.45, y = 0.55, yend = 0.55, linewidth = 0.4) +
    annotate("segment", x = 0.95, xend = 0.90, y = 0.10, yend = 0.55, linewidth = 0.4) +
    annotate("segment", x = 0.90, xend = 0.55, y = 0.55, yend = 0.55, linewidth = 0.4) +
    annotate("text", x = 0.5, y = 0.65, label = label,
             size = 3.8, fontface = "plain", vjust = 0) +
    coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE) +
    theme_void(base_family = "Arial")
}
```

```{r build-panel}
conds <- levels(dat$condition)
ages  <- levels(dat$age)

banner <- ggplot() +
  annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = 1,
           fill = "#E6E6F2", colour = "grey70", linewidth = 0.3) +
  annotate("text", x = 0.5, y = 0.5, label = CELL,
           size = 7, fontface = "plain") +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE) +
  theme_void(base_family = "Arial")

age_lbl <- function(a) {
  ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = a,
             size = 4.5, fontface = "plain") +
    coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE) +
    theme_void(base_family = "Arial")
}

pies <- list()
for (cc in conds) for (a in ages) {
  pies[[length(pies) + 1]] <- pie_one(dat |> filter(condition == cc, age == a))
}

plots <- c(
  list(banner),
  list(age_lbl("P11"), age_lbl("P45")),
  list(condition_header(conds[1])), pies[1:2],
  list(condition_header(conds[2])), pies[3:4],
  list(condition_header(conds[3])), pies[5:6],
  list(condition_header(conds[4])), pies[7:8]
)
design <- "
  AA
  BC
  DD
  EF
  GG
  HI
  JJ
  KL
  MM
  NO
"
heights <- c(0.55, 0.30, 0.30, 1, 0.30, 1, 0.30, 1, 0.30, 1)
body <- wrap_plots(plots, design = design, heights = heights)

panel <- cowplot::ggdraw(body)
```

# Figure 4N

```{r show, fig.width=5.5, fig.height=13}
panel
ggsave("Fig4N.pdf", plot = panel, width = 5.5, height = 13,
       device = cairo_pdf)
ggsave("Fig4N.png", plot = panel, width = 5.5, height = 13, dpi = 300, bg = "white")
```

# Tweaking the look

| Symptom                                  | Fix                                          |
|------------------------------------------|----------------------------------------------|
| Wedges in the wrong order                | edit `plot_order` in YAML params              |
| All pies rotated the same amount         | change `start_deg`                            |
| Wedges run the wrong way around the pie  | flip `direction` between `1` and `-1`         |
| Need a different colour for a category   | edit `cat_colors` in the `palette` chunk      |

# Session info

```{r session}
sessionInfo()
```
