---
title: "Figure 1D-E — ErasmusLadder behavioral analysis"
author: "Kratimenos Lab"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
```

# Overview

Figure 1D plots **post-perturbation step-time** (ms) across eight ErasmusLadder sessions for four experimental groups: Sal-N, Sal-Hx, LPS-N, LPS-Hx. Sessions 1–4 are training (no obstacle); sessions 5–8 are obstacle/challenge sessions.

Figure 1E plots **percent missteps** across the same eight sessions for the same four groups.

Per-mouse, per-session values are read from the consolidated dataset `Fig1_data.xlsx` (sheets: `steptimes_raw`, `steptimes_summary`, `missteps_raw`, `missteps_summary`).

Cohort: noPT+noEE subset. n: Sal-N = 10, Sal-Hx = 6, LPS-N = 6, LPS-Hx = 6.

```{r packages}
library(tidyverse)
library(readxl)
library(ggtext)
library(lme4)
library(lmerTest)
library(emmeans)
library(knitr)

DATA_FILE <- "data/Fig1_data.xlsx"

group_levels <- c("Sal-N", "Sal-Hx", "LPS-N", "LPS-Hx")
group_colors <- c(
  "Sal-N"  = "#308C59",
  "Sal-Hx" = "#4A6DE1",
  "LPS-N"  = "#FA9009",
  "LPS-Hx" = "#DA1C40"
)

theme_fig1 <- theme_classic(base_size = 11, base_family = "Arial") +
  theme(
    panel.grid.major = element_line(color = "grey92", linewidth = 0.3),
    legend.title = element_blank(),
    legend.key.size = unit(0.4, "cm"),
    legend.text = element_text(size = 9),
    legend.background = element_rect(fill = "white", color = "black", linewidth = 0.3),
    legend.position = c(0.99, 0.99),
    legend.justification = c(1, 1),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10)
  )
```

```{r load}
load_raw <- function(sheet, value_name) {
  read_excel(DATA_FILE, sheet = sheet) |>
    pivot_longer(starts_with("Session"),
                 names_to = "Session", values_to = value_name,
                 names_prefix = "Session") |>
    mutate(Session = as.integer(Session),
           Group = factor(Condition, levels = group_levels)) |>
    select(Group, Subject, Session, !!sym(value_name))
}

fig1d_raw <- load_raw("steptimes_raw", "PostPert_ms")
fig1e_raw <- load_raw("missteps_raw",  "Misstep_pct")

load_summary <- function(sheet) {
  read_excel(DATA_FILE, sheet = sheet) |>
    mutate(Group = factor(Condition, levels = group_levels),
           Phase = if_else(Session <= 4, "Training", "Obstacle"))
}
sumD <- load_summary("steptimes_summary")
sumE <- load_summary("missteps_summary")
```

# Statistics

Two-way repeated-measures ANOVA with Tukey's multiple comparisons, fit as a linear mixed-effects model (`lmer(y ~ Group * Session + (1|MouseID))`) so post-hoc contrasts use the pooled within-subject error term. Type-III ANOVA with Satterthwaite df; pairwise contrasts use Kenward–Roger df with Tukey adjustment.

```{r stats-fig1d}
fig1d_for_lmer <- fig1d_raw |>
  mutate(Session = factor(Session), MouseID = paste(Group, Subject, sep = "_"))

fit1d <- lmer(PostPert_ms ~ Group * Session + (1 | MouseID),
              data = fig1d_for_lmer, REML = TRUE)
anova(fit1d) |> as.data.frame() |>
  rownames_to_column("Effect") |>
  kable(caption = "Fig 1D — Type-III ANOVA", digits = 4)

emm1d  <- emmeans(fit1d, ~ Group | Session, lmer.df = "kenward-roger")
pwc1d  <- pairs(emm1d, adjust = "tukey") |> as.data.frame()
pwc1d_s5 <- pwc1d |> filter(Session == "5") |> arrange(p.value)
pwc1d_s5 |>
  select(contrast, estimate, SE, df, t.ratio, p.value) |>
  kable(caption = "Fig 1D — Tukey pairwise (Group | Session = 5)", digits = 4)
```

```{r stats-fig1e}
fig1e_for_lmer <- fig1e_raw |>
  mutate(Session = factor(Session), MouseID = paste(Group, Subject, sep = "_"))

fit1e <- lmer(Misstep_pct ~ Group * Session + (1 | MouseID),
              data = fig1e_for_lmer, REML = TRUE)
anova(fit1e) |> as.data.frame() |>
  rownames_to_column("Effect") |>
  kable(caption = "Fig 1E — Type-III ANOVA", digits = 4)

emm1e  <- emmeans(fit1e, ~ Group | Session, lmer.df = "kenward-roger")
pwc1e  <- pairs(emm1e, adjust = "tukey") |> as.data.frame()
pwc1e_s5 <- pwc1e |> filter(Session == "5") |> arrange(p.value)
pwc1e_s5 |>
  select(contrast, estimate, SE, df, t.ratio, p.value) |>
  kable(caption = "Fig 1E — Tukey pairwise (Group | Session = 5)", digits = 4)
```

```{r helpers}
sig_stars <- function(p) {
  case_when(
    p < 0.0001 ~ "****",
    p < 0.001  ~ "***",
    p < 0.01   ~ "**",
    p < 0.05   ~ "*",
    TRUE       ~ "ns"
  )
}

sig_caption <- function(pwc_session) {
  sig <- pwc_session |>
    filter(p.value < 0.05) |>
    arrange(p.value)
  if (nrow(sig) == 0) {
    return("<span style='color:#666666;font-style:italic'>Tukey HSD &middot; session 5</span><br><span style='color:#888888'>No significant pairwise comparisons.</span>")
  }
  # Use the HTML entity &#42; for the asterisk so Markdown doesn't interpret it
  asterisks_html <- function(p) strrep("&#42;",
                                       nchar(sig_stars(p)))
  parts <- vapply(seq_len(nrow(sig)), function(i) {
    sprintf("<span style='color:#1a1a1a;font-weight:700'>%s</span>&nbsp;<span style='color:#444444'>%s</span>",
            asterisks_html(sig$p.value[i]),
            gsub(" - ", " <span style='font-style:italic;color:#777'>vs</span> ",
                 gsub("[()]", "", as.character(sig$contrast[i]))))
  }, character(1))
  paste0(
    "<span style='color:#555555;font-style:italic;letter-spacing:0.5px'>Tukey HSD &middot; session 5</span><br>",
    paste(parts, collapse = " <span style='color:#bbbbbb'>&nbsp;&#9474;&nbsp;</span> ")
  )
}

# Dot-to-dot significance brackets, drawn on the LEFT side of session 5.
# Each bracket spans the y-range between the two compared group means.
# Short horizontal caps point right toward the data; the rotated asterisks
# sit just to the LEFT of each vertical bar. Most-significant bracket is
# closest to the data (innermost); least-significant is furthest left.
dot_to_dot_brackets <- function(sumdf, pwc_session, x_data = 5,
                                first_offset = -0.14, step = -0.26) {
  sig <- pwc_session |>
    filter(p.value < 0.05) |>
    arrange(p.value)            # most significant first → innermost
  if (nrow(sig) == 0) return(list())

  s5_means <- sumdf |> filter(Session == x_data) |>
    select(Group, Mean) |> deframe()

  layers <- list()
  for (i in seq_len(nrow(sig))) {
    contrast_str <- gsub("[()]", "", as.character(sig$contrast[i]))
    grp <- strsplit(contrast_str, " - ")[[1]]
    y1 <- s5_means[[grp[1]]]; y2 <- s5_means[[grp[2]]]
    y_bot <- min(y1, y2); y_top <- max(y1, y2)
    x_b <- x_data + first_offset + (i - 1) * step  # negative offset → bracket sits on left

    cap_w <- 0.06   # short horizontal tick at each end (does NOT reach the dots)
    layers <- c(layers, list(
      # Vertical bar
      annotate("segment", x = x_b, xend = x_b,
               y = y_bot, yend = y_top,
               linewidth = 0.5, color = "black"),
      # Bottom tick (short cap pointing toward dots)
      annotate("segment", x = x_b, xend = x_b + cap_w,
               y = y_bot, yend = y_bot,
               linewidth = 0.5, color = "black"),
      # Top tick
      annotate("segment", x = x_b, xend = x_b + cap_w,
               y = y_top, yend = y_top,
               linewidth = 0.5, color = "black"),
      # Asterisks to the LEFT of the bracket, rotated 90° (vertical text)
      annotate("text", x = x_b - 0.04, y = (y_bot + y_top) / 2,
               label = sig_stars(sig$p.value[i]),
               size = 5, hjust = 0.5, vjust = 0.5,
               angle = 90, fontface = "bold")
    ))
  }
  layers
}

sig_table <- function(pwc_session, panel_label) {
  pwc_session |>
    filter(p.value < 0.05) |>
    arrange(p.value) |>
    mutate(Significance = sig_stars(p.value),
           Comparison = gsub(" - ", " vs ",
                             gsub("[()]", "", as.character(contrast))),
           `P (Tukey)` = signif(p.value, 3)) |>
    select(Comparison, Significance, `P (Tukey)`) |>
    kable(caption = paste0(panel_label,
            " — significant pairwise comparisons at session 5 (Tukey-adjusted)"))
}

# Plot helper. Lines break between Training (1–4) and Obstacle (5–8) using
# `aes(group = interaction(Group, Phase))`; the x-axis itself stays continuous.
# The significant comparisons are added as a single-line caption below the plot.
base_panel <- function(sumdf, ylab, ymax, ybreaks, pwc_session, caption_text) {
  ggplot(sumdf, aes(x = Session, y = Mean, color = Group)) +
    geom_errorbar(aes(ymin = Mean - SEM, ymax = Mean + SEM),
                  width = 0.18, linewidth = 0.55) +
    geom_line(aes(group = interaction(Group, Phase)),
              linetype = "11", linewidth = 0.85) +
    geom_point(size = 2.8, shape = 16) +
    dot_to_dot_brackets(sumdf, pwc_session, x_data = 5) +
    annotate("text", x = 2.5, y = ymax * 0.99,
             label = "Training Sessions", size = 3.6, color = "grey25") +
    annotate("text", x = 6.5, y = ymax * 0.99,
             label = "Obstacle Sessions", size = 3.6, color = "grey25") +
    scale_color_manual(values = group_colors) +
    scale_x_continuous(breaks = 1:8, expand = expansion(mult = 0.02)) +
    scale_y_continuous(limits = c(0, ymax), breaks = ybreaks,
                       expand = expansion(mult = c(0, 0.02))) +
    labs(x = "Sessions", y = ylab, caption = caption_text) +
    theme_fig1 +
    theme(plot.caption = element_markdown(hjust = 0.5, size = 9.5,
                                          color = "grey20",
                                          lineheight = 1.4,
                                          margin = margin(t = 12)))
}
```

# Figure 1D — Post-perturbation step-time

```{r fig1d, fig.width=8, fig.height=5.2}
# In-figure Tukey caption suppressed (caption_text = NULL) to match the
# published Figure 1 layout. The post-hoc comparisons are still reported in
# the kable table below the chart for the HTML report.
p1d <- base_panel(sumD, "Post-perturbation step-time (ms)", 1000,
                  seq(0, 1000, 200),
                  pwc_session = pwc1d_s5,
                  caption_text = NULL)
p1d
ggsave("Fig1D.pdf", p1d, width = 8, height = 5.2,
       device = cairo_pdf)
ggsave("Fig1D.png", p1d, width = 8, height = 5.2, dpi = 300, bg = "white")
```

```{r fig1d-sig, results='asis'}
sig_table(pwc1d_s5, "Fig 1D")
```

# Figure 1E — % Misstep

```{r fig1e, fig.width=8, fig.height=5.2}
# In-figure Tukey caption suppressed; see fig1d chunk above for rationale.
p1e <- base_panel(sumE, "%Misstep", 25, seq(0, 25, 5),
                  pwc_session = pwc1e_s5,
                  caption_text = NULL)
p1e
ggsave("Fig1E.pdf", p1e, width = 8, height = 5.2,
       device = cairo_pdf)
ggsave("Fig1E.png", p1e, width = 8, height = 5.2, dpi = 300, bg = "white")
```

```{r fig1e-sig, results='asis'}
sig_table(pwc1e_s5, "Fig 1E")
```

# Per-group summary tables

```{r summary-tables}
sumD |>
  select(Group, Session, Mean, SEM, N) |>
  kable(caption = "Fig 1D — group means ± SEM by session", digits = 2)

sumE |>
  select(Group, Session, Mean, SEM, N) |>
  kable(caption = "Fig 1E — group means ± SEM by session", digits = 2)
```

# Session info

```{r}
sessionInfo()
```
