---
title: "Figure S4 — GSEA GO enrichment of CoGAPS patterns M-1 and M-4"
subtitle: "Per-pattern top-25 GO terms — single-column dotplots"
author: "Kratimenos Lab"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7, dpi = 300)
suppressPackageStartupMessages({
  library(readxl); library(dplyr); library(tidyr); library(tibble)
  library(ggplot2); library(patchwork); library(knitr); library(kableExtra)
})
```

# Overview

Figure S4 shows the top 25 GO terms (MSigDB C5 \ HPO; positive-tail
`fgsea` scoring) for the two CoGAPS patterns highlighted in Figure 3:

- **M-1** — Synaptic-structure / neurotransmission / membrane-signalling
  (the pattern with the strongest signed FDR in **Sal-Hx** at P45).
- **M-4** — Electron-transport chain / ATP synthesis / mitochondrial
  organisation (the pattern with the strongest signed FDR in **LPS-N**
  and **LPS-Hx** at P45).

Each panel is a single-column dotplot where the **dot size** encodes
–log₁₀(p) and the **dot color** encodes NES; pathways are sorted by
ascending p-value, most significant on top.

# Source data

Pre-computed `fgsea` results from the mouse cerebellum CoGAPS n=15
(folia-restricted) run. Each sheet (`Pattern_1` … `Pattern_15`) lists
`pathway, pval, padj, NES, size` for every GO term tested.

```{r load}
gsea_xlsx <- "data/mouse_CB_folia_cogaps_n15_gsea_fgseaGO_geneWeights.xlsx"

sheets <- c("Pattern_1", "Pattern_4")
pattern_labels <- c("Pattern_1" = "M-1", "Pattern_4" = "M-4")

df <- lapply(sheets, function(s) {
  read_excel(gsea_xlsx, sheet = s) %>%
    select(pathway, pval, padj, NES, size) %>%
    mutate(Sheet = s)
}) %>% bind_rows()

top_n <- 25
```

```{r build}
build_single_pattern <- function(focus_sheet) {
  df %>%
    filter(Sheet == focus_sheet) %>%
    arrange(pval) %>%
    slice_head(n = top_n) %>%
    mutate(
      Pattern = factor(pattern_labels[Sheet], levels = c("M-1", "M-4")),
      Pathway = factor(pathway, levels = rev(pathway)),
      nlog10p = -log10(pval)
    )
}

panel_M1 <- build_single_pattern("Pattern_1")
panel_M4 <- build_single_pattern("Pattern_4")
stopifnot(nrow(panel_M1) == top_n, nrow(panel_M4) == top_n)

# Shared scale limits so the two panels are visually comparable
nes_lim  <- range(c(panel_M1$NES,     panel_M4$NES),     na.rm = TRUE)
size_lim <- range(c(panel_M1$nlog10p, panel_M4$nlog10p), na.rm = TRUE)
```

# Per-pattern summary

```{r counts}
tibble(
  Pattern   = c("M-1", "M-4"),
  Top_N     = c(nrow(panel_M1), nrow(panel_M4)),
  Min_pval  = c(sprintf("%.2e", min(panel_M1$pval)),
                sprintf("%.2e", min(panel_M4$pval))),
  Max_NES   = c(sprintf("%.2f", max(panel_M1$NES)),
                sprintf("%.2f", max(panel_M4$NES))),
  Top_pathway = c(as.character(panel_M1$pathway[which.min(panel_M1$pval)]),
                  as.character(panel_M4$pathway[which.min(panel_M4$pval)]))
) %>%
  kable(caption = "Per-pattern summary (top 25 GO terms by p-value)") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

# Figure S4 — single-column dotplots

```{r figs4, fig.width=10, fig.height=7}
make_single_col <- function(d, title, show_legend = FALSE) {
  ggplot(d, aes(x = Pattern, y = Pathway)) +
    geom_point(aes(size = nlog10p, color = NES)) +
    scale_color_gradient(low = "#FBE6E1", high = "#B40426",
                         limits = nes_lim, name = "NES") +
    scale_size_continuous(range = c(1.5, 7), limits = size_lim,
                          name = "-Log10(P)") +
    theme_minimal(base_size = 9, base_family = "Arial") +
    theme(
      axis.text.y        = element_text(size = 7),
      axis.text.x        = element_text(size = 11, face = "bold"),
      panel.grid.minor   = element_blank(),
      panel.grid.major.x = element_blank(),
      legend.position    = if (show_legend) "right" else "none",
      plot.title         = element_text(hjust = 0.5, size = 11)
    ) +
    labs(x = NULL, y = NULL, title = title)
}

p_M1 <- make_single_col(panel_M1, "M-1 (top 25)", show_legend = FALSE)
p_M4 <- make_single_col(panel_M4, "M-4 (top 25)", show_legend = TRUE)

p_figs4 <- p_M1 + p_M4 + plot_layout(ncol = 2)
p_figs4

ggsave("FigS4.pdf", p_figs4, width = 10, height = 7,
       device = cairo_pdf)
ggsave("FigS4.png", p_figs4, width = 10, height = 7, dpi = 300, bg = "white")
```

# Session info

```{r session}
sessionInfo()
```
