---
title: "Figure 5D — p27 immunoreactivity at P11"
author: "Kratimenos Lab"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 2.7, fig.height = 5.0, dpi = 96)
suppressPackageStartupMessages({
  library(tidyverse); library(rstatix); library(ggtext); library(knitr); library(scales)
})

PANEL <- "D"
DATA_FILE <- "data/Fig5CD_data.csv"
group_levels <- c("Sal-N","Sal-Hx","LPS-N","LPS-Hx")
group_fill <- c("Sal-N"="#C6D0E2","Sal-Hx"="#7A8FA8","LPS-N"="#3A4D68","LPS-Hx"="#C8B8BC")
group_dot  <- c("Sal-N"="#4A6FA5","Sal-Hx"="#2C4A6E","LPS-N"="#0D1F38","LPS-Hx"="#A55C66")
theme_panel <- theme_classic(base_size = 13, base_family = "Arial") +
  theme(panel.grid = element_blank(), legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x  = element_text(size = 12, color = "black", angle = 90, hjust = 1, vjust = 0.5),
        plot.title = element_text(size = 11, face = "plain", color = "grey35", hjust = 0.5, margin = margin(0, 0, 6, 0)))
p_to_stars <- function(p) {
  if (is.na(p)) return("")
  if (p < 0.0001) "****" else if (p < 0.001) "***" else
  if (p < 0.01)  "**"   else if (p < 0.05) "*"   else "ns"
}
```

# Overview

**Figure 5D** — p27 immunoreactivity at P11. Per-mouse density = p27+ nuclei per mm³ of
cerebellar cortex (60x confocal, DAPI-counterstained). n = 5-6 per group.
Error bars = mean +/- SD (per the published figure legend).

Test: one-way ANOVA; post-hoc: uncorrected Fisher's LSD (6 pairwise comparisons,
alpha = 0.05). Matches the published Prism workflow exactly.

```{r load}
d <- read_csv(DATA_FILE, show_col_types = FALSE) |>
  filter(panel == PANEL) |>
  mutate(group = factor(group, levels = group_levels))
kable(d |> count(group, name = "n_mice"),
      caption = sprintf("Per-group n (per-mouse) — Panel %s", PANEL))
```

# One-way ANOVA omnibus

```{r aov}
a <- summary(aov(value ~ group, data = d))[[1]]
data.frame(test = "One-way ANOVA",
           F_stat = signif(a$`F value`[1], 4),
           df = paste(a$Df[1], a$Df[2], sep = ","),
           p_value = signif(a$`Pr(>F)`[1], 4),
           signif = p_to_stars(a$`Pr(>F)`[1])) |>
  kable(caption = "Omnibus one-way ANOVA")
```

# Uncorrected Fisher's LSD post-hoc

```{r lsd}
fisher_lsd <- function(d) {
  fit  <- aov(value ~ group, data = d)
  s    <- summary(fit)[[1]]
  mse  <- s$`Mean Sq`[2]
  df_e <- s$Df[2]
  grp  <- levels(droplevels(d$group))
  cmb  <- combn(grp, 2, simplify = FALSE)
  bind_rows(lapply(cmb, function(pair) {
    x <- d$value[d$group == pair[1]]
    y <- d$value[d$group == pair[2]]
    md <- mean(x) - mean(y)
    se <- sqrt(mse * (1/length(x) + 1/length(y)))
    tstat <- md / se
    pval  <- 2 * pt(-abs(tstat), df_e)
    tibble(group1 = pair[1], group2 = pair[2],
           mean_diff = md, se = se, t = tstat, df = df_e, p_value = pval)
  }))
}
ph <- fisher_lsd(d)
ph |>
  mutate(p_value = signif(p_value, 4),
         signif  = vapply(p_value, p_to_stars, character(1))) |>
  kable(caption = "Uncorrected Fisher's LSD pairwise contrasts.",
        digits = 4)
```

# Per-group summary

```{r summary}
d |> group_by(group) |>
  summarise(N = n(),
            Mean = round(mean(value), 2),
            SD   = round(sd(value),   2),
            SEM  = round(sd(value)/sqrt(n()), 2),
            .groups = "drop") |>
  kable(caption = "Per-group mean +/- SD (per-mouse values)")
```

# Figure 5D

```{r plot, fig.width=2.7, fig.height=5.0}
sig <- ph |>
  filter(p_value < 0.05) |>
  arrange(p_value) |>
  mutate(label = vapply(p_value, p_to_stars, character(1)),
         xmin  = match(group1, group_levels),
         xmax  = match(group2, group_levels))
ymax     <- 1e5
y_breaks <- seq(0, 1e5, by = 2e4)
ypad <- if (nrow(sig) > 0) ymax * (1 + 0.08 * (1 + nrow(sig))) else ymax * 1.05

p <- ggplot(d, aes(x = group, y = value)) +
  stat_summary(aes(fill = group), fun = mean, geom = "col",
               width = 0.65, color = "black", linewidth = 0.40, alpha = 0.60) +
  stat_summary(fun.data = function(z) {
                 m <- mean(z); s <- sd(z)
                 data.frame(y = m, ymin = m - s, ymax = m + s)
               },
               geom = "errorbar",
               width = 0.18, linewidth = 0.55, color = "black") +
  geom_jitter(aes(fill = group, color = group), shape = 21,
              width = 0.07, size = 2.0, stroke = 0.5, alpha = 0.90) +
  scale_fill_manual(values = group_fill) +
  scale_color_manual(values = group_dot) +
  scale_x_discrete(expand = expansion(add = c(0.55, 0.40))) +
  scale_y_continuous(breaks = y_breaks,
                     expand = expansion(mult = c(0, 0)),
                     labels = scales::label_number(scale = 1e-3, suffix = "k")) +
  coord_cartesian(ylim = c(0, ypad), clip = "off") +
  labs(y = "p27 (cells/mm³)", title = "p27 density (P11)") +
  theme_panel

if (nrow(sig) > 0) {
  sig$y <- ymax * (1.00 + 0.07 * seq_len(nrow(sig)))
  p <- p +
    geom_segment(data = sig, aes(x = xmin, xend = xmax, y = y, yend = y),
                 inherit.aes = FALSE, linewidth = 0.4) +
    geom_segment(data = sig, aes(x = xmin, xend = xmin,
                                 y = y, yend = y - ymax * 0.015),
                 inherit.aes = FALSE, linewidth = 0.4) +
    geom_segment(data = sig, aes(x = xmax, xend = xmax,
                                 y = y, yend = y - ymax * 0.015),
                 inherit.aes = FALSE, linewidth = 0.4) +
    geom_text(data = sig, aes(x = (xmin + xmax)/2,
                              y = y + ymax * 0.035, label = label),
              inherit.aes = FALSE, size = 4, fontface = "bold")
}
p
ggsave("Fig5D.pdf", plot = p, width = 2.7, height = 5.0,
       device = cairo_pdf)
ggsave("Fig5D.png", plot = p, width = 2.7, height = 5.0, dpi = 300, bg = "white")
```

# Session info

```{r session}
sessionInfo()
```
