---
title: "Figure 4A — Mitochondrial Proton Leak at P11"
author: "Kratimenos Lab"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 2.8, fig.height = 5.4, dpi = 96)
```

# Overview

**Figure 4A** — mitochondrial proton leak (pmol O₂/min) at **P11** (n = 9 / 14 / 8 / 9 for Sal-N / Sal-Hx / LPS-N / LPS-Hx; two P11 mice excluded as Grubbs borderline outliers — see below). Bars = group mean; error bars = SEM; dots = individual mice. No pairwise contrasts at P11 reach p < 0.05 in the combined two-way GROUP × AGE mixed-effects model (the significant contrasts at P45 are shown in Figure 4B).

```{r packages}
library(tidyverse)
library(readxl)
library(rstatix)
library(emmeans)
library(outliers)
library(ggtext)
library(knitr)

DATA_FILE <- "data/Fig4ABC_data.xlsx"
group_levels <- c("Sal-N", "Sal-Hx", "LPS-N", "LPS-Hx")
group_fill   <- c("Sal-N"="#C6D0E2", "Sal-Hx"="#7A8FA8",
                  "LPS-N"="#3A4D68", "LPS-Hx"="#C8B8BC")
group_dot    <- c("Sal-N"="#4A6FA5", "Sal-Hx"="#2C4A6E",
                  "LPS-N"="#0D1F38", "LPS-Hx"="#A55C66")
theme_panel <- theme_classic(base_size = 13, base_family = "Arial") +
  theme(
    panel.grid = element_blank(), legend.position = "none",
    axis.title.y = element_text(size = 11, margin = margin(r = 5)),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size = 10, color = "black"),
    axis.text.x  = element_text(size = 13, color = "black",
                                angle = 90, hjust = 1, vjust = 0.5),
    axis.line  = element_line(linewidth = 0.5),
    axis.ticks = element_line(linewidth = 0.5),
    plot.title = element_textbox_simple(
      size = 10, hjust = 0.5, halign = 0.5,
      fill = "#E8E5EE", color = "grey15", box.color = NA,
      padding = margin(4, 8, 4, 8), margin = margin(0, 0, 6, 0),
      r = unit(2, "pt"), width = grid::unit(0.65, "npc")),
    plot.margin = margin(6, 10, 6, 6)
  )
```

```{r load}
d_full <- bind_rows(
  read_excel(DATA_FILE, sheet = "protonleak_p11") |> mutate(Age = "P11"),
  read_excel(DATA_FILE, sheet = "protonleak_p45") |> mutate(Age = "P45")
) |> mutate(Group = factor(Group, levels = group_levels),
            Age   = factor(Age,   levels = c("P11","P45"))) |>
  rename(y = ProtonLeak_pmol_min)
```

# Outlier analysis (P11)

Grubbs' test (one-sided, most-extreme high value) per P11 group; borderline threshold p < 0.10.

```{r grubbs}
grubbs_p11 <- d_full |> filter(Age == "P11") |>
  group_by(Group) |>
  summarise(N = n(),
            Max_value   = round(max(y), 2),
            Max_mouse   = MouseID[which.max(y)],
            G_statistic = round(grubbs.test(y, type = 10)$statistic[[1]], 3),
            P_value     = round(grubbs.test(y, type = 10)$p.value, 4),
            .groups = "drop") |>
  mutate(Decision = ifelse(P_value < 0.10, "Exclude (borderline outlier)", "Retain"))
kable(grubbs_p11, caption = "Grubbs test for a single high outlier — P11 (one-sided)")
```

- **P11 LPS-Hx**: mouse **P11-DH10 = 91.95 pmol/min** (G = 2.14, p = 0.062) — excluded.
- **P11 LPS-N**: mouse **P11-LPSNx9 = 61.76 pmol/min** (G = 2.10, p = 0.053) — excluded.

```{r exclude}
d <- d_full |> filter(!(Age == "P11" & MouseID %in% c("P11-DH10", "P11-LPSNx9")))
d |> filter(Age == "P11") |> count(Group) |>
  kable(caption = "P11 cohort after outlier exclusion")
```

# Combined statistics (P11 + P45)

The asterisks shown in Figure 4B come from a single two-way GROUP × AGE model
fit to both timepoints. We run that full model here so the P11 stats are
reported in their proper context.

```{r aov-2way}
aov_2way <- aov(y ~ Group * Age, data = d)
broom::tidy(aov_2way) |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  kable(caption = "Two-way ANOVA — Proton Leak ~ Group × Age (n = 63)")
```

```{r lsmeans}
emm <- emmeans(aov_2way, ~ Group | Age)
pw  <- as.data.frame(pairs(emm, adjust = "none"))
sig_stars <- function(p) case_when(p < 0.0001 ~ "****", p < 0.001 ~ "***",
                                    p < 0.01 ~ "**", p < 0.05 ~ "*", TRUE ~ "ns")
pw |>
  mutate(across(where(is.numeric), \(x) round(x, 4)),
         signif = sig_stars(p.value)) |>
  filter(Age == "P11") |>
  kable(caption = "LSMEANS pairwise contrasts at P11 (unadjusted). None reach p < 0.05.")
```

# Per-group summary — P11

```{r summary-tab}
d |> filter(Age == "P11") |>
  group_by(Group) |>
  summarise(N = n(), Mean = round(mean(y), 2),
            SEM  = round(sd(y) / sqrt(n()), 2), .groups = "drop") |>
  kable(caption = "P11 group mean ± SEM after outlier exclusion")
```

# Figure 4A

```{r fig4a, fig.width=2.8, fig.height=5.4, fig.cap="Figure 4A. Mitochondrial proton leak at P11 (pmol O₂/min). Bars = group mean; error bars = SEM; dots = individual mice. n = 9/14/8/9. Two P11 outliers excluded (P11-DH10 LPS-Hx, p = 0.062; P11-LPSNx9 LPS-N, p = 0.053)."}
p4a <- ggplot(d |> filter(Age == "P11"), aes(x = Group, y = y)) +
  stat_summary(aes(fill = Group), fun = mean, geom = "col",
               width = 0.45, color = "black", linewidth = 0.40, alpha = 0.60) +
  stat_summary(fun.data = mean_se, geom = "errorbar",
               width = 0.14, linewidth = 0.50, color = "black") +
  geom_jitter(aes(fill = Group, color = Group), shape = 21,
              width = 0.09, size = 2.4, stroke = 0.6) +
  scale_fill_manual(values = group_fill) +
  scale_color_manual(values = group_dot) +
  scale_x_discrete(expand = expansion(add = c(0.55, 0.30))) +
  scale_y_continuous(limits = c(0, 130),
                     breaks = c(0, 20, 40, 60, 80, 100),
                     expand = expansion(mult = c(0, 0))) +
  labs(y = expression("OCR (pmol O"[2]*"/min)"),
       title = "P11 Proton leak") +
  theme_panel
p4a
ggsave("Fig4A.pdf", plot = p4a, width = 2.6, height = 5.4,
       device = cairo_pdf)
ggsave("Fig4A.png", plot = p4a, width = 4.5, height = 5.4, dpi = 300, bg = "white")
```

# Session info

```{r session}
sessionInfo()
```
