---
title: "Figure 4B — Mitochondrial Proton Leak at P45"
author: "Kratimenos Lab"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 2.8, fig.height = 5.4, dpi = 96)
```

# Overview

**Figure 4B** — mitochondrial proton leak (pmol O₂/min) at **P45** (n = 6 / 5 / 6 / 6 for Sal-N / Sal-Hx / LPS-N / LPS-Hx). Bars = group mean; error bars = SEM; dots = individual mice. Asterisks mark pairwise LSMEANS contrasts at p < 0.05 from the combined two-way GROUP × AGE mixed-effects model (the model is fit on both P11 + P45 data).

```{r packages}
library(tidyverse)
library(readxl)
library(rstatix)
library(emmeans)
library(outliers)
library(ggtext)
library(knitr)

DATA_FILE <- "data/Fig4ABC_data.xlsx"
group_levels <- c("Sal-N", "Sal-Hx", "LPS-N", "LPS-Hx")
group_fill   <- c("Sal-N"="#C6D0E2", "Sal-Hx"="#7A8FA8",
                  "LPS-N"="#3A4D68", "LPS-Hx"="#C8B8BC")
group_dot    <- c("Sal-N"="#4A6FA5", "Sal-Hx"="#2C4A6E",
                  "LPS-N"="#0D1F38", "LPS-Hx"="#A55C66")
theme_panel <- theme_classic(base_size = 13, base_family = "Arial") +
  theme(
    panel.grid = element_blank(), legend.position = "none",
    axis.title.y = element_text(size = 11, margin = margin(r = 5)),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size = 10, color = "black"),
    axis.text.x  = element_text(size = 13, color = "black",
                                angle = 90, hjust = 1, vjust = 0.5),
    axis.line  = element_line(linewidth = 0.5),
    axis.ticks = element_line(linewidth = 0.5),
    plot.title = element_textbox_simple(
      size = 10, hjust = 0.5, halign = 0.5,
      fill = "#E8E5EE", color = "grey15", box.color = NA,
      padding = margin(4, 8, 4, 8), margin = margin(0, 0, 6, 0),
      r = unit(2, "pt"), width = grid::unit(0.65, "npc")),
    plot.margin = margin(6, 10, 6, 6)
  )
```

```{r load}
d_full <- bind_rows(
  read_excel(DATA_FILE, sheet = "protonleak_p11") |> mutate(Age = "P11"),
  read_excel(DATA_FILE, sheet = "protonleak_p45") |> mutate(Age = "P45")
) |> mutate(Group = factor(Group, levels = group_levels),
            Age   = factor(Age,   levels = c("P11","P45"))) |>
  rename(y = ProtonLeak_pmol_min)
# Apply the same Grubbs-based outlier exclusion as Fig4A
d <- d_full |> filter(!(Age == "P11" & MouseID %in% c("P11-DH10", "P11-LPSNx9")))
```

# Combined statistics (P11 + P45)

The two asterisks on Panel B come from a single two-way GROUP × AGE model
fit on both timepoints; reported here so they're regenerated from data.

```{r aov-2way}
aov_2way <- aov(y ~ Group * Age, data = d)
broom::tidy(aov_2way) |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  kable(caption = "Two-way ANOVA — Proton Leak ~ Group × Age (n = 63)")
```

```{r lsmeans}
emm <- emmeans(aov_2way, ~ Group | Age)
pw  <- as.data.frame(pairs(emm, adjust = "none"))
sig_stars <- function(p) case_when(p < 0.0001 ~ "****", p < 0.001 ~ "***",
                                    p < 0.01 ~ "**", p < 0.05 ~ "*", TRUE ~ "ns")
pw |>
  mutate(across(where(is.numeric), \(x) round(x, 4)),
         signif = sig_stars(p.value)) |>
  filter(Age == "P45") |>
  kable(caption = "LSMEANS pairwise contrasts at P45 (unadjusted)")
```

Significant contrasts at P45 (asterisks on this panel):

- **LPS-Hx vs Sal-Hx (P45):** \* p = 0.031
- **LPS-Hx vs Sal-N (P45):** \* p = 0.046

# Per-group summary — P45

```{r summary-tab}
d |> filter(Age == "P45") |>
  group_by(Group) |>
  summarise(N = n(), Mean = round(mean(y), 2),
            SEM  = round(sd(y) / sqrt(n()), 2), .groups = "drop") |>
  kable(caption = "P45 group mean ± SEM")
```

# Figure 4B

```{r fig4b, fig.width=2.8, fig.height=5.4, fig.cap="Figure 4B. Mitochondrial proton leak at P45 (pmol O₂/min). Bars = group mean; error bars = SEM; dots = individual mice. n = 6/5/6/6. Asterisks: LPS-Hx vs Sal-Hx (* p = 0.031) and LPS-Hx vs Sal-N (* p = 0.046) from the combined two-way GROUP × AGE model with pairwise LSMEANS contrasts within each Age."}

# Significance brackets — same as the original Panel B
sig_b <- list(
  annotate("segment", x = 2, xend = 4, y = 113, yend = 113,
           linewidth = 0.5, color = "black"),
  annotate("segment", x = 2, xend = 2, y = 111, yend = 113,
           linewidth = 0.5, color = "black"),
  annotate("segment", x = 4, xend = 4, y = 111, yend = 113,
           linewidth = 0.5, color = "black"),
  annotate("text",    x = 3, y = 116, label = "*",
           size = 5, fontface = "bold", color = "black"),
  annotate("segment", x = 1, xend = 4, y = 124, yend = 124,
           linewidth = 0.5, color = "black"),
  annotate("segment", x = 1, xend = 1, y = 122, yend = 124,
           linewidth = 0.5, color = "black"),
  annotate("segment", x = 4, xend = 4, y = 122, yend = 124,
           linewidth = 0.5, color = "black"),
  annotate("text",    x = 2.5, y = 127, label = "*",
           size = 5, fontface = "bold", color = "black")
)

p4b <- ggplot(d |> filter(Age == "P45"), aes(x = Group, y = y)) +
  stat_summary(aes(fill = Group), fun = mean, geom = "col",
               width = 0.45, color = "black", linewidth = 0.40, alpha = 0.60) +
  stat_summary(fun.data = mean_se, geom = "errorbar",
               width = 0.14, linewidth = 0.50, color = "black") +
  geom_jitter(aes(fill = Group, color = Group), shape = 21,
              width = 0.09, size = 2.4, stroke = 0.6) +
  scale_fill_manual(values = group_fill) +
  scale_color_manual(values = group_dot) +
  scale_x_discrete(expand = expansion(add = c(0.55, 0.30))) +
  scale_y_continuous(limits = c(0, 130),
                     breaks = c(0, 20, 40, 60, 80, 100),
                     expand = expansion(mult = c(0, 0))) +
  labs(y = expression("OCR (pmol O"[2]*"/min)"),
       title = "P45 Proton leak") +
  theme_panel
for (lyr in sig_b) p4b <- p4b + lyr
p4b
ggsave("Fig4B.pdf", plot = p4b, width = 2.6, height = 5.4,
       device = cairo_pdf)
ggsave("Fig4B.png", plot = p4b, width = 4.5, height = 5.4, dpi = 300, bg = "white")
```

# Session info

```{r session}
sessionInfo()
```
