---
title: "Figure 6C — VGLUT2⁺ laminar distribution (Outer/Inner ML ratio)"
author: "Kratimenos Lab"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    theme: cosmo
    toc: true
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 3.2, fig.height = 5.2, dpi = 150)
suppressPackageStartupMessages({
  library(tidyverse); library(FSA); library(knitr); library(kableExtra)
})
set.seed(42)
```

# Overview

Per-animal **Outer/Inner ML VGLUT2⁺ puncta-density ratio** for the four
perinatal-insult groups (n = 6 / 4 / 5 / 5 for Sal-N / Sal-Hx / LPS-N /
LPS-Hx). A ratio of 1 (dashed line) indicates equal puncta density across
the inner and outer halves of the molecular layer.

Statistical test: **Kruskal–Wallis** (primary, non-parametric) with Dunn
pairwise post-hoc, BH-adjusted across 6 pairs. One-way ANOVA shown for
reference.

# Data

```{r data}
project_dir <- "."
derived     <- file.path(project_dir, "derived")

group_levels <- c("Sal-N","Sal-Hx","LPS-N","LPS-Hx")
group_fill   <- c("Sal-N"="#C6D0E2","Sal-Hx"="#7A8FA8",
                  "LPS-N"="#3A4D68","LPS-Hx"="#C8B8BC")

per_band <- read_csv(file.path(derived, "per_animal_per_band.csv"),
                     show_col_types = FALSE) |>
  mutate(group_label = factor(group_label, levels = group_levels))

per_band |>
  group_by(group_label) |>
  summarise(n = n(),
            `Inner density (per μm²)` = signif(mean(inner_density_per_um2), 4),
            `Outer density (per μm²)` = signif(mean(outer_density_per_um2), 4),
            `Outer/Inner ratio`       = signif(mean(outer_inner_ratio),     3),
            .groups = "drop") |>
  kable(caption = "Per-group descriptives.") |>
  kable_styling(full_width = FALSE)
```

# Statistical tests

```{r stats}
p_label <- function(p) {
  if (is.na(p))         ""
  else if (p < 0.0001)  "****"
  else if (p < 0.001)   "***"
  else if (p < 0.01)    "**"
  else if (p < 0.05)    "*"
  else                  ""
}

s_kw  <- kruskal.test(outer_inner_ratio ~ group_label, data = per_band)
s_anv <- aov(outer_inner_ratio ~ group_label, data = per_band)
s_anv_s <- summary(s_anv)[[1]]

tibble(
  Test = c("Kruskal–Wallis (primary)",
           "One-way ANOVA (sensitivity)"),
  Statistic = c(sprintf("chi² = %.3f", s_kw$statistic),
                sprintf("F = %.3f",    s_anv_s$`F value`[1])),
  df = c(as.character(s_kw$parameter),
         sprintf("%d, %d", s_anv_s$Df[1], s_anv_s$Df[2])),
  `p-value` = signif(c(s_kw$p.value, s_anv_s$`Pr(>F)`[1]), 4)
) |>
  kable(caption = "Omnibus tests.") |>
  kable_styling(full_width = FALSE)

# Dunn pairwise — robust to group label hyphens (Sal-Hx etc.) by parsing
# group occurrences anywhere in the comparison string.
parse_pair_positions <- function(pair_str, levels) {
  s <- gsub("\\s+", "", pair_str)
  hits <- vapply(levels, \(g) grepl(paste0("(^|-)", g, "(-|$)"), s),
                 logical(1))
  m <- levels[hits]
  if (length(m) != 2) return(c(NA_integer_, NA_integer_))
  sort(match(m, levels))
}

dunn_raw <- FSA::dunnTest(outer_inner_ratio ~ group_label,
                          data = per_band, method = "none")$res |>
  as_tibble()
positions <- t(sapply(dunn_raw$Comparison, parse_pair_positions,
                      levels = group_levels))
dunn <- dunn_raw |>
  mutate(g_lo  = group_levels[positions[, 1]],
         g_hi  = group_levels[positions[, 2]],
         p_adj = p.adjust(P.unadj, method = "BH")) |>
  transmute(Contrast = paste(g_lo, "vs", g_hi),
            group1 = g_lo, group2 = g_hi,
            `Dunn Z`   = signif(Z, 3),
            `p (raw)`  = signif(P.unadj, 4),
            `p (BH-6)` = signif(p_adj, 4),
            Sig = vapply(p_adj, p_label, character(1)),
            p_adj = p_adj)

dunn |> select(-group1, -group2, -p_adj) |>
  kable(caption = "Dunn pairwise (BH across 6 pairs).") |>
  kable_styling(full_width = FALSE)
```

# Figure

```{r plot, fig.width=3.2, fig.height=5.2}
ymax_s <- max(per_band$outer_inner_ratio, na.rm = TRUE)
top_s  <- max(1.2, ymax_s * 1.6)

p_ratio <- ggplot(per_band, aes(x = group_label, y = outer_inner_ratio)) +
  geom_boxplot(aes(fill = group_label), alpha = 0.55,
               width = 0.40, outlier.shape = NA, color = "black",
               linewidth = 0.35) +
  geom_jitter(aes(fill = group_label), shape = 21, width = 0.08,
              size = 2.4, stroke = 0.5) +
  geom_hline(yintercept = 1, linetype = 2, color = "grey50") +
  scale_fill_manual(values = group_fill) +
  scale_x_discrete(expand = expansion(add = c(0.55, 0.30))) +
  coord_cartesian(ylim = c(0, top_s)) +
  labs(x = NULL, y = "Outer / Inner puncta-density ratio",
       title = expression("VGLUT2"^"+"~"laminar distribution")) +
  theme_classic(base_size = 13, base_family = "Arial") +
  theme(legend.position = "none",
        plot.title.position = "plot",
        plot.title   = element_text(size = 12, face = "bold",
                                    hjust = 0.5, margin = margin(b = 6)),
        axis.title.y = element_text(size = 13, color = "black",
                                    margin = margin(r = 5)),
        axis.text.x  = element_text(color = "black", size = 12,
                                    angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y  = element_text(color = "black", size = 12),
        plot.margin  = margin(8, 8, 4, 4))

# Significance brackets for significant Dunn pairs
sig <- dunn |> filter(!is.na(p_adj), p_adj < 0.05) |> arrange(p_adj)
if (nrow(sig) > 0) {
  y_base <- ymax_s * 1.10; step <- ymax_s * 0.13
  for (i in seq_len(nrow(sig))) {
    x1 <- match(sig$group1[i], group_levels)
    x2 <- match(sig$group2[i], group_levels)
    yb <- y_base + (i - 1) * step
    p_ratio <- p_ratio +
      annotate("segment", x = min(x1,x2), xend = max(x1,x2),
               y = yb, yend = yb, linewidth = 0.5) +
      annotate("segment", x = min(x1,x2), xend = min(x1,x2),
               y = yb, yend = yb - 0.02 * ymax_s, linewidth = 0.5) +
      annotate("segment", x = max(x1,x2), xend = max(x1,x2),
               y = yb, yend = yb - 0.02 * ymax_s, linewidth = 0.5) +
      annotate("text", x = (x1 + x2) / 2, y = yb + 0.025 * ymax_s,
               label = p_label(sig$p_adj[i]), size = 5, fontface = "bold")
  }
}

ggsave("Fig6C.pdf", p_ratio, width = 3.2, height = 5.2, device = cairo_pdf)
ggsave("Fig6C.png", p_ratio, width = 3.2, height = 5.2, dpi = 600, bg = "white")
p_ratio
```

# Session info

```{r session, echo=FALSE}
sessionInfo()
```
